Modeling trend in temperature volatility using generalized LASSO

### 2018-05-18 nips-2018/paper.tex
page length ok, to arash for final edits

\begin{abstract}
words, words,

\end{abstract}

@@

\attn{Arash: Some equations do not have numbering and some have.}

\attn{Arash: fix appendicies and the references to them.}

Nonparametric variance estimation for spatio-temporal data.

\subsection{Motivating applications}

\attn{cut this down}

There is a considerable interest in determining if there is an
increasing trend in the climate variability
\citep{hansen_perception_2012,huntingford_no_2013}. An increase in the
temperature variability will increase the probability of extreme hot
outliers. It might be harder for the society to adapt to these
extremes than to the gradual increase in the mean temperature
\citep{huntingford_no_2013}.

In this paper, we consider the problem of detecting the trend in the
temperature volatility. All analyses are performed on a sub-set of the
European Centre for Medium-Range Weather Forecasts (ECMWF) ERA-20C
dataset \citep{uppala_era-40_2005}. This dataset include the
temperature measurements over a grid over the earth from 1957 to
2002.~\citep{VasseurDeLong2014,TrenberthZhang2014,StatenKahn2016,Screen2014,FischerBeyerle2013}

Research on analyzing the trend in the volatility of spatio-temporal
data is scarce. \citep{hansen_perception_2012} studied the change in
the standard deviation (SD) of the surface temperature in the NASA
Goddard Institute for Space Studies gridded temperature data set. In
their analysis, for each geographical position, the mean of the
temperature computed for the period 1951-1980 (called the base-period)
at that position, is subtracted from the corresponding time
series. Each time series is then divided by the standard deviation
computed at each position and during the same time period. The
distribution of the resulting data is then plotted for different
periods. These distributions represent the deviation of the
temperature for a specific period, from the mean in the base period,
in units of the standard deviation in that period. The results showed
that these distributions are widen for the resent time periods
compared to 1951-1980.  \citep{huntingford_no_2013} took a similar
approach in analysing the ERA-40 data set. However, in addition to the
aforementioned method, they computed the distribution of the SDs in an
alternative way: for each position and each time period, the deviation
of the time-series at that position from the mean in that time period
at that position was computed, and then divided by the SD of that
position in the period before 1981. The results showed that there
still is an increase in the SDs from 1958-1970 to 1991-2001, but this
is much less than what is obtained from the method used in
\citep{hansen_perception_2012}. The authors also computed the
time-evolving global SD from the de-trended time-series at each
position. The resulting curve suggested that the global SD has been
stable.

These previous work (and other related research, e.g.,
\citep{rhines_frequent_2013}) have several shortcomings. First, no
statistical analysis has been performed to examine if the change in
the SD is statistically significant. Second, the methodologies for
computing the SDs are rather arbitrary. The deviation of each
time-series in a given period, is computed from either the mean of a
base-period (as in \citep{hansen_perception_2012}), or from the given
period (as in \citep{rhines_frequent_2013,huntingford_no_2013}). These
deviations are then normalized using the SD of the base-period or the
given period. No justification is provided for these choices. Third,
the correlation between the observations is ignored. The observations
in subsequent days and close geographical positions could be highly
correlated. Without considering these correlations, any conclusion
based on the averaged data could be flawed.

The main contribution of this work is to develop a new methodology for
detecting the trend in the volatility of spatio-temporal data. In this
methodology, the variance at each position and time, is considered as
a hidden (un-observed) variable. The value of these hidden variables
are then estimated by maximizing the likelihood of the observed
data. We show that this formulation per se, is not appropriate for
detecting the trend. To overcome this issue, we penalize the
differences between the estimated variances of the observations which
are temporally and/or spatially close to each other. This will result
in an optimization problem called the \textit{generalized LASSO
  problem} \citep{tibshirani_solution_2011}. As we will see, the
dimension of this optimization problem is very high and so the
standard methods for solving the generalized LASSO cannot be applied
directly. We investigate two methods for solving this optimization
problem. In the first method, we adopt an optimization technique
called alternative direction method of multipliers (ADMM)
\citep{boyd_distributed_2011}, to divide the total problem into
several sub-problems of much lower dimension and show how the total
problem can be solved by iteratively solving these sub-problems. The
second method, called the \textit{linearized ADMM algorithm}
\citep{parikh_proximal_2014} solves the main problem by iteratively
solving a linearized version of it. We will compare the benefits of
each method.

Also neuroscience.

@@

Mention ~\citep{KimKoh2009,HallacPark2017}. Also,
~\citep{Tibshirani2014,TibshiraniTaylor2011}. ARCH/GARCH.

\citep{SadhanalaWang2017,LinSharpnack2017,WangSharpnack2016}~\citep{RamdasTibshirani2016}

@@

\begin{itemize}
\item We propose a model for non-parametric variance estimation for a
  spatio-temporal process (~\autoref{sec:ell_1-trend-filt}).
\item We derive two algorithms to fit our estimator when applied to
  very large data (~\autoref{sec:prop-optim-meth}).
\item We illustrate our methods on a large global temperature dataset
  with the goal of tracking world-wide trends in variance as well as a
  simulation constructed to mimic these data's features
  (~\autoref{sec:empirical-evaluation}). 
\end{itemize}

@@

$\ell_1$-trend filtering was proposed by \citep{KimKoh2009} as a method for estimating a smooth, time-varying trend. It is formulated as the optimization problem
% : a time-series $y_t$, $t=1,...,T$ is observed and we believe that
% it consists of a smooth trend and a stochastic component and the
% goal is to estimate the trend. The estimated trend is desired to be
% smooth, but at the same time, we seek small estimated residuals
% (stochastic component). These two objectives cannot be achieved
% simultaneously and so a trade-off between them should be made. In
% $\ell_1$-trend filtering, this trade-off is formulated as the
% following convex optimization problem: 
\begin{equation}
\min_{\beta} \frac{1}{2} \sum_{t=1}^{T} (y_t-\beta_t)^2+\lambda
\sum_{t=1}^{T-2} \left|\beta_t-2\beta_{t+1}+\beta_{t+2}\right| 
\end{equation}
or equivalently:
\begin{equation}
\min_{\beta} \frac{1}{2} \norm{ y-\beta }_2^2+\lambda \norm{ D\beta}_1
\label{eq:l1tf}
\end{equation}
 where $y_t$ is an observed time-series, $\beta$ is the smooth trend,
 $D$ is a $(T-2)\times T$ matrix, and $\lambda$ is a tuning parameter
 which balances fidelity to the data (small errors in the first term)
 with a desire for smoothness.  
% The \textit{penalty matrix} is:
% \begin{equation}
% D=\begin{bmatrix}
% 1 & -2 & 1 &  &  &  &\\ 
% & 1 & -2 & 1 &  &  &\\ 
% &  & \ddots & \ddots & \ddots  &  &\\ 
% &  & & 1 & -2 & 1 &  \\ 
% &  &  &   & 1 & -2 & 1 
% \end{bmatrix}
% \label{eq:d_matrix}
% \end{equation}
%The first term in \eqref{eq:l1tf} penalizes large residuals while the second term encourages smooth estimated $\beta$.
With the penalty matrix $D$, the estimated $\beta$ will be piecewise
linear. \citep{KimKoh2009} proposed a specialized primal-dual
interior point (PDIP) algorithm for solving \eqref{eq:l1tf}. From a
statistical perspective, \eqref{eq:l1tf} is a constrained maximum
likelihood problem with independent observations from a normal
distribution with common variance, $y_t \sim \mbox{N}(\beta_t,
\sigma^2)$, subject to a piecewise linear constraint on $\beta$.

@@

The first term in the objective is proportional to the negative
log-likelihood, the second is the temporal penalty for the
time-series at each location $(i,j)$, while the third and fourth,
penalize the difference between the estimated variance of two
vertically and horizontally adjacent points, respectively. The spatial
component of this
penalty is a special case of trend
filtering on graphs~\citep{WangSharpnack2016} which penalizes the difference
between the estimated values of the signal on the connected
nodes. As before, we can write
\eqref{eq:l1tf_var_st} in matrix form where $h$ is an $T\times
S$ vector and $D$ is replaced by $D_{TS} \in \mathbb{R}^{(N_t+N_s)
  \times (T \cdot S)	}$, where $N_t=S \cdot (T-2)$ and $N_s=T \cdot
(2n_rn_c-n_r)$ are the number of temporal and spatial constraints,
respectively\footnote{$N_s$ is obtained by counting the number of
  unique constraints at each location and at all times.}. Then, as we 
have two different tuning parameters for the temporal and spatial
components, we write $\Lambda =\left[\lambda_1\one_{N_t}^\top,\;
  \lambda_2\one_{N_s}^\top\right]^\top$ leading
to:\footnote{Throughout the paper, we use $|x|$ for both scalars and
  vectors. For vectors we use this to denote a vector obtained by
  taking the absolute value of each entry of $x$.} 
% let $h$ be a
% vector whose first $T$ entries are $h_{11t}$ for $t=1,...,T$, its next
% $T$ entries are $h_{21t}$ and so on, then the optimization problem in
% the matrix form is as
\begin{equation}
\min_h -l(y\given h)+ \Lambda^\top | D_{TS}h |.
\label{eq:l1tf_var_st_mat}
\end{equation}
\attn{equations should have proper punctuation as if they were a sentence.}
% where $\Lambda = \mathbf{I}_{Tn_rn_c}\left[\lambda_1\one_T^\top,\;
%   \lambda_2\one_{n_rn_c}^\top\right]$ and 
% $D_{total}^t=[D_{t}^t|D_{s}^t]$, where $n_t$ and $n_s$ are the number
% of rows of $D_{t}$ and $D_{s}$, respectively (see below). The matrix
% $D_t$ is the following block-diagonal matrix and corresponds to the
% temporal penalty: 
% \begin{equation}
% D_t=\begin{bmatrix}
% D &  & \\ 
%  & \ddots & \\
%  &  & D
% \end{bmatrix}
% \label{eq:d_t_matrix}
% \end{equation}
%  where $D$ was defined in \eqref{eq:d_matrix}. The number of the
%  diagonal blocks is equal to the grid size $n_r \times n_c$. Each row
%  of the matrix $D_s$ corresponds to one spatial constraint in
%  \eqref{eq:l1tf_var_st}. For example, the first $T$ rows correspond to
%  $|h_{11t}-h_{21t}|$ for $t=1,...,T$, the next $T$ rows correspond to
%  $|h_{11t}-h_{12t}|$, and so on. 
% \attn{Detailed discussion of the PDIP to here}

@@

Assume it is possible to define a set of local variables
$x_i \in \mathbb{R}^{n_i}$ such that $f(h)=\sum_i f_i(x_i)$, where
each $x_i$ is a subset of the global variable $h$. More specifically,
each entry of the local variables corresponds to an entry of the
global variable. Therefore we can define a mapping $\mathscr{G}(i,j)$
from the local variable indices into the global variable indices:
$k=\mathscr{G}(i,j)$ means that the $j^\text{th}$ entry of $x_i$ is
$h_k$ (or $(x_i)_j=h_k$). For ease of notation, define $\tilde{h}_i
\in \mathbb{R}^{n_i}$ as $(\tilde{h}_i)_j=h_{\mathscr{G}(i,j)}$. Then,
the original optimization problem is equivalent to the following
problem:  
\begin{equation}
\begin{aligned}
\min_{\{x_1,...,x_N  \}} \sum_i f_i(x_i)\\
 s.t. \,\,\, \tilde{h}_i=x_i. \,\,\,\,\,\,
\end{aligned}
\label{eq:consADMM}
\end{equation}
It is important to note that each entry of the global variable may
correspond to several entries of the local variables and so the
constraints $\tilde{h}_i=x_i$ enforce the consensus between the local
variables corresponding to the same global variable.

The augmented Lagrangian corresponding to 
\eqref{eq:consADMM} is $L_\rho(x,h,y)=\sum_i
\big(f_i(x_i)+u_i^\top(x_i-\tilde{h}_i) + (\rho/2) \lVert
x_i-\tilde{h}_i \lVert_2^2 \big)$. Now, we can apply ADMM to $L_\rho$
which results in the following ADMM updates:
\begin{align}
x_i&\leftarrow\argmin_{x_i} f_i(x_i) + (u_i)^\top x_i +
     (\rho/2)  \norm{ x_i-\tilde{h}_i }_2^2\\
h_k&\leftarrow (1/S_k)\sum_{\mathscr{G}(i,j)=k} (x_i)_j \label{eq:ADMM_steps}\\
u_i&\leftarrow u_i + \rho (x_i-\tilde{h}_i).
\end{align}
\attn{If you hate the left arrow, that's fine. But change it
  everywhere, and use $=$. The $:=$ notation means that ``thing on
  left is defined to be thing on right'' and isn't really appropriate here.}
Here, $S_k$ is the number of local variable entries that correspond to
$h_k$, and $u_i$ are the Lagrange multipliers.

To solve the optimization problem \eqref{eq:l1tf_var_st_mat} using
this method, we need to address two 
questions: first, how to choose the 
local variables $x_i$, and second, how to the update in the first line of
\eqref{eq:ADMM_steps}.

\begin{figure}[tb]
  \centering
  \includegraphics[height=.2\textheight]{Figures/data_cube}
  \caption{The cube represents the global variable $h$ in space and
    time. The sub-cubes specified by the white lines are
    $x_i$.}
  \label{fig:data_cube}
\end{figure}

In \autoref{fig:data_cube}, the global variable $h$ is represented
as a cube (using the subset of the US as an example). We decompose $h$
into sub-cubes as shown by white lines. With
this definition of $x_i$, the objective
\eqref{eq:l1tf_var_st_mat} decomposes as $\sum_i f_i(x_i)$ where
$f_i(x_i)=-l(y_i\given x_i)+\Lambda_{(i)}^\top |D_{(i)}x_i|$, and
$\Lambda_{(i)}$ and $D_{(i)}$ contain the temporal and spatial
penalties corresponding to $x_i$ only. 
% The $x$-update step in
% \eqref{eq:ADMM_steps} is the following optimization problem:
% $x_i^{m+1}:=\argmin_{x_i} ( f_i(x_i) + 
% (u_i^m)^\top x_i + (\rho/2) \lVert x_i-\tilde{z}_i^m \lVert_2^2)$.
Thus, with this choice of the local variables $x_i$, we solve the
$x$-update using the PDIP method. 
% As it 
% was explained in \autoref{sec:l1tf_var}, to use PDIP we need to
% compute the dual problem, which in turn needs the computation of the
% Fenchel conjugate of the loss function. In addition, referring to
% \autoref{alg:pdip}, we need to compute the gradient and Jacobian of
% the conjugate functions. It is important to note that compare to the
% optimization problem \autoref{eq:l1tf_var}, the loss function in the
% x-update step in \autoref{eq:ADMM_steps} includes the quadratic term
% $(\rho/2) \lVert x_i-\tilde{z}_i^m \lVert_2^2$. This makes the
% computation more involved than \autoref{sec:l1tf_var}. 
The details of the remaining computations
the computations are explained in the Supplement.

% The complete ADMM algorithm for estimating the variances is
% represented in \autoref{alg:ADMM}. All the computations in the three
% updating steps \eqref{eq:ADMM_steps} can be performed in parallel. The
% number of rows and columns of the sub-cubes should be chosen so that
% the updating of $x_i$ could be performed in one processor. We choose
% $3\times3\times521$ sub-cubes.

% \begin{algorithm}[tb]
%   \caption{ADMM for sparse estimation of variance of spatio-temporal
%     data \attn{Fix this}}
%   \label{alg:ADMM}
%   \begin{algorithmic}
%     \STATE {\bfseries Input:} data $y$, mapping $\mathscr{G}(i,j)$,
%     $\rho,\lambda_t,\lambda_s$ 
%     \STATE Initialization: $x_i^0=z^0=u_i^0=\textbf{0}$.
%     \FOR{$m=1,2,...$ }
%     \FOR{$i=1$ {\bfseries to} $N_{sub-cubes}$}
%     \STATE compute $\nu_i$ from \eqref{eq:dual_x_update}
%     \STATE compute $w_i$ from \eqref{eq:conj_x_update}
%     \STATE set $x_i^m:=w_i$
%     \ENDFOR
%     \STATE Compute $z^m$ from  \eqref{eq:ADMM_steps}
%     \STATE Compute $u_i^m$ from  \eqref{eq:ADMM_steps}
%     \ENDFOR
%   \end{algorithmic}
% \end{algorithm}

@@

Consider the generic optimization problem
$\min_x f(x)+g(Dx)$
where $x\in \mathbb{R}^n$ and $D\in \mathbb{R}^{m\times n}$. Each
iteration of the linearized ADMM
algorithm~\citep{parikh_proximal_2014} for solving this problem 
has the form
\begin{align}
x & \leftarrow \prox_{\mu f} \left(x - (\mu/\rho)D^\top (D x - z + u )\right)\\
z & \leftarrow \prox_{\rho g} \left(z + u\right) \label{eq:linADMM_steps}\\
u & \leftarrow u + D x - z
\end{align}
where the algorithm parameters $\mu$ and $\rho$ satisfy $0 < \mu < \rho/\norm{D}_2^2$, $z,u\in \mathbb{R}^m$ and the proximal operator is defined as
\begin{equation}
 \prox_{\alpha f}(u) = \min_x \,\, \alpha \cdot f(x)+\frac{1}{2} \norm{ x-u}_2^2.
\label{eq:linADMM_prox}
\end{equation}

@@

In words, the variance at each time and location is computed as the
weighted sum of $S$ bell-shaped functions where the weights are
time-varying, consist of a linear trend $\alpha_s \cdot t$ and a
periodic term $\beta_s \cdot \sin(2\pi\omega_s t+\phi_s)$. The
bell-shaped functions impose the spatial smoothness, and the linear
trend and the periodic terms enforce the temporal smoothness similar
to the seasonal component in the real climate data. We simulated the
data on a 5 by 7 grid and for 780 time steps with $S=4$. 
\attn{The
parameters of the variance function are shown in
\autoref{tab:sim_params} in the Supplement. For reference, we plot the
variance function for all locations at $t=25$ and $t=45$ in as well as
the variance across time at $(0,0)$ in \autoref{fig:true_var_spatial}
in Appendix C.}

@@

\paragraph{Analysis of trend of temperature volatility}

>>>>

\begin{abstract}
In this paper, we present methodology for estimating trends in
spatio-temporal volatility. We give two algorithms for computing our
estimator which are tailored for dense, gridded observations over both
space and time, though these can be easily extended to other
structures (time-varying network flows, neuroimaging). We motivate our
methodology using by applying it to a massive climate dataset and
discuss the implications for climate analysis.
\end{abstract}

@@

Estimating smooth trends over time for large collections of time
series is a common problem in economics, finance, meteorology,
neuroscience and more. Most of this work focuses on analyzing trends
(or removing trends) in the temporal average, but trends in variance
can be more relevant, especially for financial data but also for
climate science.

Trends in terrestrial temperature variability are perhaps more relavant for species
viability than trends in mean
temperature~\citep{huntingford_no_2013},
because, an increase in the
temperature variability will increase the probability of extreme hot
outliers~\citep{VasseurDeLong2014}. Recent climate literature
suggests that it is more difficult for society to adapt to these
extremes than to the gradual increase in the mean temperature
\citep{hansen_perception_2012,huntingford_no_2013}. Furthermore, the willingness of popular media to
emphasize the prevalence extreme cold events coupled with a
fundamental misunderstanding of the relationship between climate (the global
distribution of weather over the long run) and weather (observed
short-term, localized behavior) leads to public misunderstanding of climate
change. In fact, increased frequency of extreme cold events in the
northern hemisphere is can be partially attributed to increases in
mean climate but is also due to increases in temperature
variance~\citep{Screen2014,FischerBeyerle2013,TrenberthZhang2014}.

Nevertheless, research examining trends in the volatility of
spatio-temporal climate data is scarce. \citep{hansen_perception_2012} studied the change in
the standard deviation (SD) of the surface temperature in the NASA
Goddard Institute for Space Studies gridded temperature data set by
examining the empirical SD at each spatial location relative to that
location's SD over a base period, and showed
that these estimates are increasing.
\citep{huntingford_no_2013} took a similar
approach in analyzing the ERA-40 data set. Their results showed that there
still is an increase in the SDs from 1958-1970 to 1991-2001, but this
is much less than what is obtained from the method used in
\citep{hansen_perception_2012}. The authors also computed the
time-evolving global SD from the de-trended time-series at each
position, which suggests that the global SD has been
stable.

These and other related research, e.g.,
\citep{rhines_frequent_2013}) have several shortcomings. First, no
statistical analysis has been performed to examine if the changes in
the SD are statistically significant. Second, the methodologies for
computing the SDs are highly sensitive to the choice of base period. Third,
and most importantly, temporal and spatial correlations between the
observations are ignored.

@@

Variance estimation for financial time series has a lengthy history,
focused especially on parametric models like the generalized
autoregressive conditional heteroskedasticity (GARCH) process~\citep{engle2002dynamic} and
stochastic volatility models~\citep{HarveyRuiz1994}. These models (and
related AR processes) are specifically for parametric modelling of
short ``bursts'' of high volatility, behavior typical of financial
instruments. Parametric models for spatial data go back at least
to~\citep{besag1974spatial} who proposed a conditional probability
model on the lattice for examining plant ecology.

More recently, nonparametric models for both spatial and temporal data
have focused on using $\ell_1$-regularization for trend
estimation. \citep{KimKoh2009} proposed $\ell_1$-trend filtering for
univariate time series, which forms the basis of our methods. These
methods have been generalized to higher order temporal smoothness
~\citep{Tibshirani2014}, graph dependencies~\citep{WangSharpnack2016},
and, most recently, small, time-varying
graphs~\citep{HallacPark2017}. Our methodology is similar in flavor
to~\citep{HallacPark2017}, though it uses a different likelihood
function to emphasize variance estimation rather than trends in mean
signal. Furthermore, the focus is in high-dimensional, regular data
rather than a small number of changing graphs.

@@

The main contribution of this work is to develop a new methodology for
detecting the trend in the volatility of spatio-temporal data. In this
methodology, the variance at each position and time is considered as
a hidden variable. The values of these hidden variables
are then estimated by maximizing the likelihood of the observed
data. We show that this formulation is not appropriate for
detecting the trend, so, following ~\citep{Tibshirani2014}, we penalize the
differences between the estimated variances which
are temporally and spatially ``close'', resulting
in a generalized LASSO problem. % However, the
% dimension of this optimization problem is very high and so the
% standard methods solvers are inadequate. We develop two algorithms
% which are computationally feasible. In the first method, we adopt an optimization technique
% called alternative direction method of multipliers (ADMM)
% \citep{boyd_distributed_2011}, to divide the total problem into
% several sub-problems of much lower dimension and show how the total
% problem can be solved by iteratively solving these sub-problems. The
% second method, called the \textit{linearized ADMM algorithm}
% \citep{parikh_proximal_2014} solves the main problem by iteratively
% solving a linearized version of it. We will compare the benefits of
% each method.

Our main contributions are as follows:
\begin{enumerate}
\item We propose a model for nonparametric variance estimation for a
  spatio-temporal process (\autoref{sec:ell_1-trend-filt}).
\item We derive two alternating direction method of multiplier
  algorithms (ADMM) to fit our estimator when applied to 
  very large data (\autoref{sec:prop-optim-meth}). We give situations
  under which each algorithm is most likely to be useful.
\item We illustrate our methods on a large global temperature dataset
  with the goal of tracking world-wide trends in variance as well as a
  simulation constructed to mimic these data's features
  (\autoref{sec:empirical-evaluation}). 
\end{enumerate}

While we motivate and illustrate our methods on large, gridded climate
data, we note that our algorithms are also applicable to neuroimaging
data or large collections of financial instruments.

@@

$\ell_1$-trend filtering was proposed by \citep{KimKoh2009} as a
method for estimating a smooth, time-varying trend. It is formulated
as the optimization problem 
% : a time-series $y_t$, $t=1,...,T$ is observed and we believe that
% it consists of a smooth trend and a stochastic component and the
% goal is to estimate the trend. The estimated trend is desired to be
% smooth, but at the same time, we seek small estimated residuals
% (stochastic component). These two objectives cannot be achieved
% simultaneously and so a trade-off between them should be made. In
% $\ell_1$-trend filtering, this trade-off is formulated as the
% following convex optimization problem: 
$
\min_{\beta} \frac{1}{2} \sum_{t=1}^{T} (y_t-\beta_t)^2+\lambda
\sum_{t=1}^{T-2} \left|\beta_t-2\beta_{t+1}+\beta_{t+2}\right| 
$
or equivalently:
\begin{equation}
\min_{\beta} \frac{1}{2} \norm{ y-\beta }_2^2+\lambda \norm{ D\beta}_1
\label{eq:l1tf}
\end{equation}
 where $y_t$ is an observed time-series, $\beta$ is the smooth trend,
 $D$ is a $(T-2)\times T$ matrix, and $\lambda$ is a tuning parameter
 which balances fidelity to the data (small errors in the first term)
 with a desire for smoothness.  
% The \textit{penalty matrix} is:
% \begin{equation}
% D=\begin{bmatrix}
% 1 & -2 & 1 &  &  &  &\\ 
% & 1 & -2 & 1 &  &  &\\ 
% &  & \ddots & \ddots & \ddots  &  &\\ 
% &  & & 1 & -2 & 1 &  \\ 
% &  &  &   & 1 & -2 & 1 
% \end{bmatrix}
% \label{eq:d_matrix}
% \end{equation}
%The first term in \eqref{eq:l1tf} penalizes large residuals while the second term encourages smooth estimated $\beta$.
With the penalty matrix $D$, the estimated $\beta$ will be piecewise
linear. \citep{KimKoh2009} proposed a specialized primal-dual
interior point (PDIP) algorithm for solving \eqref{eq:l1tf}. From a
statistical perspective, \eqref{eq:l1tf} is a constrained maximum
likelihood problem with independent observations from a normal
distribution with common variance, $y_t \sim \mbox{N}(\beta_t,
\sigma^2)$, subject to a piecewise linear constraint on $\beta$.

@@

The first term in the objective is proportional to the negative
log-likelihood, the second is the temporal penalty for the
time-series at each location $(i,j)$, while the third and fourth,
penalize the difference between the estimated variance of two
vertically and horizontally adjacent points, respectively. The spatial
component of this
penalty is a special case of trend
filtering on graphs~\citep{WangSharpnack2016} which penalizes the difference
between the estimated values of the signal on the connected
nodes. As before, we can write
\eqref{eq:l1tf_var_st} in matrix form where $h$ is an $T\times
S$ vector and $D$ is replaced by $D_{ST} \in \mathbb{R}^{(N_t+N_s)
  \times (T \cdot S)	}$, where $N_t=S \cdot (T-2)$ and $N_s=T \cdot
(2n_rn_c-n_r)$ are the number of temporal and spatial constraints,
respectively\footnote{$N_s$ is obtained by counting the number of
  unique constraints at each location and at all times.}. Then, as we 
have two different tuning parameters for the temporal and spatial
components, we write $\Lambda =\left[\lambda_1\one_{N_t}^\top,\;
  \lambda_2\one_{N_s}^\top\right]^\top$ leading
to:\footnote{Throughout the paper, we use $|x|$ for both scalars and
  vectors. For vectors we use this to denote a vector obtained by
  taking the absolute value of each entry of $x$.} 
% let $h$ be a
% vector whose first $T$ entries are $h_{11t}$ for $t=1,...,T$, its next
% $T$ entries are $h_{21t}$ and so on, then the optimization problem in
% the matrix form is as
\begin{equation}
\min_h -l(y\given h)+ \Lambda^\top | D_{ST}h |.
\label{eq:l1tf_var_st_mat}
\end{equation}
\attn{equations should have proper punctuation as if they were a sentence.}
% where $\Lambda = \mathbf{I}_{Tn_rn_c}\left[\lambda_1\one_T^\top,\;
%   \lambda_2\one_{n_rn_c}^\top\right]$ and 
% $D_{total}^t=[D_{t}^t|D_{s}^t]$, where $n_t$ and $n_s$ are the number
% of rows of $D_{t}$ and $D_{s}$, respectively (see below). The matrix
% $D_t$ is the following block-diagonal matrix and corresponds to the
% temporal penalty: 
% \begin{equation}
% D_t=\begin{bmatrix}
% D &  & \\ 
%  & \ddots & \\
%  &  & D
% \end{bmatrix}
% \label{eq:d_t_matrix}
% \end{equation}
%  where $D$ was defined in \eqref{eq:d_matrix}. The number of the
%  diagonal blocks is equal to the grid size $n_r \times n_c$. Each row
%  of the matrix $D_s$ corresponds to one spatial constraint in
%  \eqref{eq:l1tf_var_st}. For example, the first $T$ rows correspond to
%  $|h_{11t}-h_{21t}|$ for $t=1,...,T$, the next $T$ rows correspond to
%  $|h_{11t}-h_{12t}|$, and so on. 
% \attn{Detailed discussion of the PDIP to here}

@@

Assume it is possible to define a set of local variables
$x_i \in \mathbb{R}^{n_i}$ such that $f(h)=\sum_i f_i(x_i)$, where
each $x_i$ is a subset of the global variable $h$. More specifically,
each entry of the local variables corresponds to an entry of the
global variable. Therefore we can define a mapping $\mathscr{G}(i,j)$
from the local variable indices into the global variable indices:
$k=\mathscr{G}(i,j)$ means that the $j^\text{th}$ entry of $x_i$ is
$h_k$ (or $(x_i)_j=h_k$). For ease of notation, define $\tilde{h}_i
\in \mathbb{R}^{n_i}$ as $(\tilde{h}_i)_j=h_{\mathscr{G}(i,j)}$. Then,
the original optimization problem is equivalent to the following
problem:  
\begin{equation}
\begin{aligned}
\min_{\{x_1,...,x_N  \}} \sum_i f_i(x_i)\\
 s.t. \,\,\, \tilde{h}_i=x_i. \,\,\,\,\,\,
\end{aligned}
\label{eq:consADMM}
\end{equation}
It is important to note that each entry of the global variable may
correspond to several entries of the local variables and so the
constraints $\tilde{h}_i=x_i$ enforce the consensus between the local
variables corresponding to the same global variable.  
The augmented Lagrangian corresponding to 
\eqref{eq:consADMM} is $L_\rho(x,h,y)=\sum_i
\big(f_i(x_i)+u_i^\top(x_i-\tilde{h}_i) + (\rho/2) \lVert
x_i-\tilde{h}_i \lVert_2^2 \big)$. Now, we can apply ADMM to $L_\rho$.
% which results in the following ADMM updates:
% \begin{align}
% x_i&\leftarrow\argmin_{x_i} f_i(x_i) + (u_i)^\top x_i +
%      (\rho/2)  \norm{ x_i-\tilde{h}_i }_2^2\\
% h_k&\leftarrow (1/S_k)\sum_{\mathscr{G}(i,j)=k} (x_i)_j \label{eq:ADMM_steps}\\
% u_i&\leftarrow u_i + \rho (x_i-\tilde{h}_i).
% \end{align}
% \attn{If you hate the left arrow, that's fine. But change it
%   everywhere, and use $=$. The $:=$ notation means that ``thing on
%   left is defined to be thing on right'' and isn't really appropriate here.}
% Here, $S_k$ is the number of local variable entries that correspond to
% $h_k$, and $u_i$ are the Lagrange multipliers.  
To solve the optimization problem \eqref{eq:l1tf_var_st_mat} using
this method, we need to address two 
questions: first, how to choose the 
local variables $x_i$, and second, how to the update them.
\begin{figure}[tb]
  \centering
  \includegraphics[height=.2\textheight]{Figures/data_cube}
  \caption{The cube represents the global variable $h$ in space and
    time. The sub-cubes specified by the white lines are
    $x_i$.}
  \label{fig:data_cube}
\end{figure}

In \autoref{fig:data_cube}, the global variable $h$ is represented
as a cube (using the subset of the US as an example). We decompose $h$
into sub-cubes as shown by white lines. With
this definition of $x_i$, the objective
\eqref{eq:l1tf_var_st_mat} decomposes as $\sum_i f_i(x_i)$ where
$f_i(x_i)=-l(y_i\given x_i)+\Lambda_{(i)}^\top |D_{(i)}x_i|$, and
$\Lambda_{(i)}$ and $D_{(i)}$ contain the temporal and spatial
penalties corresponding to $x_i$ only. 
% The $x$-update step in
% \eqref{eq:ADMM_steps} is the following optimization problem:
% $x_i^{m+1}:=\argmin_{x_i} ( f_i(x_i) + 
% (u_i^m)^\top x_i + (\rho/2) \lVert x_i-\tilde{z}_i^m \lVert_2^2)$.
Thus, with this choice of the local variables $x_i$, we solve the
$x$-update using the PDIP method. 
% As it 
% was explained in \autoref{sec:l1tf_var}, to use PDIP we need to
% compute the dual problem, which in turn needs the computation of the
% Fenchel conjugate of the loss function. In addition, referring to
% \autoref{alg:pdip}, we need to compute the gradient and Jacobian of
% the conjugate functions. It is important to note that compare to the
% optimization problem \autoref{eq:l1tf_var}, the loss function in the
% x-update step in \autoref{eq:ADMM_steps} includes the quadratic term
% $(\rho/2) \lVert x_i-\tilde{z}_i^m \lVert_2^2$. This makes the
% computation more involved than \autoref{sec:l1tf_var}. 
\autoref{alg:conADMM} gives the general version of this algorithm. A
more detailed discussion of the is in the Supplement.

\begin{algorithm}[tb]
  \caption{Consensus ADMM }
  \label{alg:conADMM}
  \begin{algorithmic}
    \STATE {\bfseries Input:} data $y$, penalty matrix $D$, 
    $\epsilon, \rho,\lambda_t,\lambda_s >0$.
    \STATE {\bf Set:} $h\leftarrow 0$, $z\leftarrow 0$, $u\leftarrow
    0$. \COMMENT{Initialization} 
    \REPEAT
    \STATE $x_i\leftarrow\argmin_{x_i} -l(y_i\given
    x_i)+\Lambda_{(i)}^\top |D_{(i)}x_i|$
    \STATE $\quad\quad\quad+ (u_i)^\top x_i +
    (\rho/2)  \lVert x_i-\tilde{h}_i \rVert_2^2$. \COMMENT{Update local
      vars using PDIP}
    \STATE $h_k\leftarrow (1/S_k)\sum_{\mathscr{G}(i,j)=k} (x_i)_j
    $. \COMMENT{Global update.}
    \STATE $ u_i\leftarrow u_i + \rho (x_i-\tilde{h}_i)$. \COMMENT{Dual update}
    \UNTIL {$\norm{h^{m+1}-h^m}_2^2\ < \epsilon$}
    \STATE {\bf Return:} $h$.
  \end{algorithmic}
\end{algorithm}

@@

Consider the generic optimization problem
$\min_x f(x)+g(Dx)$
where $x\in \mathbb{R}^n$ and $D\in \mathbb{R}^{m\times n}$. Each
iteration of the linearized ADMM
algorithm~\citep{parikh_proximal_2014} for solving this problem 
has the form
\begin{align}
x & \leftarrow \prox_{\mu f} \left(x - (\mu/\rho)D^\top (D x - z + u )\right)\\
z & \leftarrow \prox_{\rho g} \left(z + u\right) \label{eq:linADMM_steps}\\
u & \leftarrow u + D x - z
\end{align}
where the algorithm parameters $\mu$ and $\rho$ satisfy $0 < \mu < \rho/\norm{D}_2^2$, $z,u\in \mathbb{R}^m$ and the proximal operator is defined as
$
 \prox_{\alpha f}(u) = \min_x \,\, \alpha \cdot f(x)+\frac{1}{2} \norm{ x-u}_2^2.
$

@@

In words, the variance at each time and location is computed as the
weighted sum of $S$ bell-shaped functions where the weights are
time-varying, consist of a linear trend $\alpha_s \cdot t$ and a
periodic term $\beta_s \cdot \sin(2\pi\omega_s t+\phi_s)$. The
bell-shaped functions impose the spatial smoothness, and the linear
trend and the periodic terms enforce the temporal smoothness similar
to the seasonal component in the real climate data. We simulated the
data on a 5 by 7 grid and for 780 time steps with $S=4$. 
Specific parameter choices of the variance function are shown in
\autoref{tab:sim_params} in the Supplement. For illustration, we also plot the
variance function for all locations at $t=25$ and $t=45$ in as well as
the variance across time at $(0,0)$ in \autoref{fig:true_var_spatial}
in the Supplement.

@@

\paragraph{Analysis of trend in temperature volatility}
